---
title: 'integrated differential ATACseq and RNAseq'
date: today
engine: knitr
execute:
    message: false
    warning: false
format:
    html:
        self-contained: true
        toc: true
        code-fold: true
params:
  cpus: 1
  artifact_dir: NULL
  input_dir: NULL
  meta: NULL
  atac_rna_tsv:   "output/rowbind_logfc/concat.atac_rna.tsv"
  atac_rna_venn:  "output/join_atac_rna/joined_atac_rna.log"
  set_counts_tsv: "output/rowbind_count/concat.set_counts.tsv"
---

```{r setup}
#| message: false
library(broom)
library(dplyr)
library(ggplot2)
library(ggVennDiagram)
library(glue)
library(knitr)
library(plotly)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tidyr)
theme_sovacool <- ggplot2::theme_bw() +
    ggplot2::theme(
      legend.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      legend.box.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      panel.spacing = unit(1, 'lines'),
      plot.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      strip.background = element_blank(),
      strip.placement = 'outside'
    )
theme_set(theme_sovacool)
```


## Differential ATAC & RNAseq 

### Overlapping genes

Differential ATACseq peaks were intersected with known gene promoter regions.
Peaks that overlap promoter regions were kept and joined with the corresponding
differential gene expression data from RNAseq.


Inspired by Fig. 4A from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9581296/figure/F4/>

```{r plot_venn}
#| message: false
v <- read_delim(params$atac_rna_venn, delim = ':\t', 
           col_names = c('set','genes')) %>% 
    pivot_wider(names_from = set, values_from = genes) %>% 
    unlist()

venn_dat <- list(
    RNAseq = identity_transformer(glue("1:{v[['RNAseq']]}")),
    ATACseq = identity_transformer(glue("{v[['RNAseq']]-v[['joined']]}:{v[['RNAseq']]+v[['ATACseq']]+v[['joined']]}"))
)

Venn(venn_dat) %>% 
    process_data(shape_id = "201") %>% 
    plot_venn() +
    scale_fill_distiller(palette = 'YlGnBu', direction = 1) +
    theme(legend.position = 'none')

```


Inspired by Fig. 3B-C from <https://www.science.org/doi/10.1126/science.abe1505#F3>

```{r plot_histogram}
overlap_counts <- read_tsv(params$set_counts_tsv) 

plot_hist <- function(dat) {
    dat %>%
        ggplot(aes(count, fill = count_type)) +
        geom_histogram(binwidth = 1) +
        facet_grid(cluster_id ~ count_type, switch = 'x') + #, strip.position = 'bottom') +
        labs(x = '') +
        theme(legend.position = 'none')
}


overlap_counts %>% 
    plot_hist() +
    scale_y_continuous(labels = label_comma())

overlap_counts %>% 
    plot_hist() +
    scale_y_log10(labels = label_log(digits = 2))

overlap_counts %>%
    filter(count > 0) %>% 
    plot_hist()
```

### Correlation

Inspiration

- Fig. 4B from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9581296/figure/F4/>
- Fig. 2A-B from <https://epigeneticsandchromatin.biomedcentral.com/articles/10.1186/s13072-019-0260-2/figures/2>


```{r read}
#| message: false
diff_dat <- read_tsv(params$atac_rna_tsv) %>% 
    mutate(cluster_id = as.factor(cluster_id))
```


```{r corr_all_genes}
alpha_level <- 0.05
log2fc_thresh <- 0 # setting to zero removes filtering by log2FC

cor_logfc <- diff_dat %>% 
    do(tidy(
        cor.test(.$log2FoldChange_atac, .$log2FoldChange_rna,
                 method = "spearman")
    ))
kable(cor_logfc)

abs_max <- 5.004292 # from filtered data. this will exclude some points.
    
diff_dat %>%
    mutate(across(starts_with('log2FoldChange'), ~ round(., 2))) %>%
    mutate(signif = case_when(
        pvalue_atac < alpha_level & pvalue_rna < alpha_level ~ TRUE,
        TRUE ~ FALSE)
    ) %>%
    dplyr::arrange(signif) %>% # plot true on top of false
    ggplot(
        aes(
            y = log2FoldChange_atac,
            x = log2FoldChange_rna
        )
    ) +
    geom_hline(
        yintercept = 0,
        linetype = 'dashed',
        color = 'darkgrey'
    ) +
    geom_vline(
        xintercept = 0,
        linetype = 'dashed',
        color = 'darkgrey'
    ) +
    geom_point(aes(fill = signif), alpha = 0.7, shape = 21, stroke = 0) +
    geom_smooth(method = lm, 
                linewidth = 0.5,
                se = FALSE, show.legend = FALSE) +
    scale_fill_manual(values = c(`TRUE` = "#252525", `FALSE` = "#D9D9D9")) +
    guides(fill = guide_legend(title = glue('ATAC & RNA\np-values < {alpha_level}'))) +
    lims(x = c(-abs_max, abs_max),
         y = c(-abs_max, abs_max)) +
    labs(y = 'ATAC log2(FC)', x = 'RNA log2(FC)')
```

### filtered to p-value < `r alpha_level`

```{r corr_signif_genes}

plot_corr_line <- function(dat, plot_lm = FALSE) {
    
    min_max <- list(min = ~ min(.x, na.rm = TRUE),
                    max = ~ max(.x, na.rm = TRUE))
    abs_max <- dat %>%
        summarize(across(starts_with('log2FoldChange'), min_max)) %>%
        as.list() %>% unname() %>% unlist() %>% abs() %>% max()
    corr_line_plot <- dat %>%
        mutate(across(starts_with('log2FoldChange'), ~ round(., 2))) %>%
        ggplot(
            aes(
                y = log2FoldChange_atac,
                x = log2FoldChange_rna,
                color = cluster_id,
                group = cluster_id,
                text = glue(
                    'gene name:\t{gene_name}',
                    'cluster:\t{cluster_id}',
                    'ATAC log2(FC):\t{log2FoldChange_atac}',
                    'RNA log2(FC):\t{log2FoldChange_rna}',
                    .sep = '\n'
                )
            )
        ) +
        geom_hline(
            yintercept = 0,
            linetype = 'dashed',
            color = 'darkgrey'
        ) +
        geom_vline(
            xintercept = 0,
            linetype = 'dashed',
            color = 'darkgrey'
        ) +
        geom_point(alpha = 0.5) +
        {if (isTRUE(plot_lm)) geom_smooth(linewidth = 0.5, method=lm, se=FALSE) } +
        scale_color_viridis_d() +
        guides(color = guide_legend(title = 'cluster')) +
        lims(x = c(-abs_max, abs_max),
             y = c(-abs_max, abs_max)) +
        labs(y = 'ATAC log2(FC)', x = 'RNA log2(FC)')
    return(corr_line_plot)
}


diff_dat_filt <- diff_dat %>% 
  filter(pvalue_rna < alpha_level, pvalue_atac < alpha_level, 
         abs(log2FoldChange_atac) >= log2fc_thresh, abs(log2FoldChange_rna) >= log2fc_thresh)

cor_logfc_filt <- diff_dat_filt %>% 
    group_by(cluster_id) %>% 
    do(tidy(
        cor.test(.$log2FoldChange_atac, .$log2FoldChange_rna,
                 method = "spearman")
    ))
cor_logfc_filt
#kable(cor_logfc)


diff_dat_filt %>% 
    select(cluster_id, starts_with('log2FoldChange_')) %>% 
    group_by(cluster_id) %>% 
    nest(data = c(log2FoldChange_atac, log2FoldChange_rna)) %>% 
    mutate(corr = map(data, ~ cor.test(.$log2FoldChange_atac, 
                                       .$log2FoldChange_rna,
                                       method = "spearman")),
           corr_tidied = map(corr, tidy)
           ) %>% 
    unnest(cols = c(corr_tidied))

filter_and_correlate <- function(df,cluster){
  df %>% filter(cluster_id == cluster) -> df_filtered
  ct <- cor.test(df_filtered$log2FoldChange_rna,df_filtered$log2FoldChange_atac,method="spearman")
  return(ct)
}
ct1 <- filter_and_correlate(diff_dat_filt,"c1")
ct1
ct2 <- filter_and_correlate(diff_dat_filt,"c2")
ct2
ct3 <- filter_and_correlate(diff_dat_filt,"c3")
ct3

lm_logfc <- diff_dat_filt %>% 
    group_by(cluster_id) %>% 
    do(tidy(lm(
        log2FoldChange_atac ~ log2FoldChange_rna, .
    )))


ggplotly(
  diff_dat_filt %>% plot_corr_line(plot_lm = TRUE), 
  tooltip = 'text'
 )
```



## Infer TF-gene regulatory network


