---
title: 'integrated differential ATACseq and RNAseq'
date: today
engine: knitr
execute:
    message: false
    warning: false
format:
    html:
        self-contained: true
        toc: true
        code-fold: true
params:
  cpus: 1
  artifact_dir: NULL
  input_dir: NULL
  meta: NULL
  atac_rna_tsv:   "output/rowbind_logfc/concat.atac_rna.tsv"
  atac_rna_venn:  "output/join_atac_rna/joined_atac_rna.log"
  set_counts_tsv: "output/rowbind_count/concat.set_counts.tsv"
  atac_rna_signif_genes_csv: 'atac_rna_signif_genes.csv'
---

```{r setup}
#| message: false
library(broom)
library(dplyr)
library(ggplot2)
library(ggVennDiagram)
library(glue)
library(knitr)
library(patchwork)
library(plotly)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tidyr)
theme_sovacool <- ggplot2::theme_bw() +
    ggplot2::theme(
      legend.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      legend.box.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      panel.spacing = unit(1, 'lines'),
      plot.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      strip.background = element_blank(),
      strip.placement = 'outside'
    )
theme_set(theme_sovacool)
```


## Differential ATAC & RNAseq 

### Overlapping genes

Differential ATACseq peaks were intersected with known gene promoter regions.
Peaks that overlap promoter regions were kept and joined with the corresponding
differential gene expression data from RNAseq.


Inspired by Fig. 4A from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9581296/figure/F4/>


```{r plot_venn}
#| message: false
v <- read_delim(params$atac_rna_venn, delim = ':\t', 
           col_names = c('set','genes')) %>% 
    pivot_wider(names_from = set, values_from = genes) %>% 
    unlist()

venn_dat <- list(
    RNAseq = identity_transformer(glue("1:{v[['RNAseq']]}")),
    ATACseq = identity_transformer(glue("{v[['RNAseq']]-v[['joined']]}:{v[['RNAseq']]+v[['ATACseq']]+v[['joined']]}"))
)

Venn(venn_dat) %>% 
    process_data(shape_id = "201") %>% 
    plot_venn() +
    scale_fill_distiller(palette = 'YlGnBu', direction = 1) +
    theme(legend.position = 'none')

```


Inspired by Fig. 3B-C from <https://www.science.org/doi/10.1126/science.abe1505#F3>

```{r plot_histogram}
overlap_counts <- read_tsv(params$set_counts_tsv) 

plot_hist <- function(dat) {
    dat %>%
        ggplot(aes(count, fill = count_type)) +
        geom_histogram(binwidth = 1) +
        facet_grid(cluster_id ~ count_type, switch = 'x') + #, strip.position = 'bottom') +
        labs(x = '') +
        theme(legend.position = 'none')
}


overlap_counts %>% 
    plot_hist() +
    scale_y_continuous(labels = label_comma())

overlap_counts %>% 
    plot_hist() +
    scale_y_log10(labels = label_log(digits = 2))

overlap_counts %>%
    filter(count > 0) %>% 
    plot_hist()
```

### Correlation

Inspiration

- Fig. 4B from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9581296/figure/F4/>
- Fig. 2A-B from <https://epigeneticsandchromatin.biomedcentral.com/articles/10.1186/s13072-019-0260-2/figures/2>


```{r read}
#| message: false
diff_dat <- read_tsv(params$atac_rna_tsv) %>% 
    mutate(cluster_id = as.factor(cluster_id))
```


```{r corr_all_genes}
alpha_level <- 0.05
log2fc_thresh <- 0 # setting to zero removes filtering by log2FC

cor_logfc <-  cor.test(diff_dat$log2FoldChange_atac, 
         diff_dat$log2FoldChange_rna,
         method = "spearman") %>% 
    tidy()
cor_logfc %>% kable()

lm_logfc <- diff_dat %>%
    lm(log2FoldChange_atac ~ log2FoldChange_rna, .) 
lm_rsq <- summary(lm_logfc)$r.squared

abs_max <- 5.004292 # from filtered data below. this will exclude some points from the unfiltered data.
    
diff_dat %>%
    mutate(across(starts_with('log2FoldChange'), ~ round(., 2))) %>%
    mutate(signif = case_when(
        pvalue_atac < alpha_level & pvalue_rna < alpha_level ~ TRUE,
        TRUE ~ FALSE)
    ) %>%
    dplyr::arrange(signif) %>% # plot true on top of false
    ggplot(
        aes(
            y = log2FoldChange_atac,
            x = log2FoldChange_rna
        )
    ) +
    geom_hline(
        yintercept = 0,
        linetype = 'dashed',
        color = 'darkgrey'
    ) +
    geom_vline(
        xintercept = 0,
        linetype = 'dashed',
        color = 'darkgrey'
    ) +
    geom_point(aes(fill = signif), alpha = 0.7, shape = 21, stroke = 0) +
    geom_smooth(method = lm, 
                linewidth = 0.5,
                se = FALSE, show.legend = FALSE) +
    scale_fill_manual(values = c(`TRUE` = "#252525", `FALSE` = "#D9D9D9")) +
    guides(fill = guide_legend(title = glue('ATAC & RNA\np-values < {alpha_level}'))) +
    lims(x = c(-abs_max, abs_max),
         y = c(-abs_max, abs_max)) +
    labs(y = 'ATAC log2(FC)', x = 'RNA log2(FC)')
```

R<sup>2</sup> = `r lm_rsq`

### filtered to p-value < `r alpha_level`

```{r corr_signif_genes}

min_max <- list(min = ~ min(.x, na.rm = TRUE),
                max = ~ max(.x, na.rm = TRUE))
plot_corr_line <- function(dat, plot_lm = FALSE) {
    
    abs_max <- dat %>%
        summarize(across(starts_with('log2FoldChange'), min_max)) %>%
        as.list() %>% unname() %>% unlist() %>% abs() %>% max()
    corr_line_plot <- dat %>%
        mutate(across(starts_with('log2FoldChange'), ~ round(., 2))) %>%
        ggplot(
            aes(
                y = log2FoldChange_atac,
                x = log2FoldChange_rna,
                color = cluster_id,
                group = cluster_id,
                text = glue(
                    'gene name:\t{gene_name}',
                    'cluster:\t{cluster_id}',
                    'ATAC log2(FC):\t{log2FoldChange_atac}',
                    'RNA log2(FC):\t{log2FoldChange_rna}',
                    .sep = '\n'
                )
            )
        ) +
        geom_hline(
            yintercept = 0,
            linetype = 'dashed',
            color = 'darkgrey'
        ) +
        geom_vline(
            xintercept = 0,
            linetype = 'dashed',
            color = 'darkgrey'
        ) +
        geom_point(alpha = 0.5) +
        {if (isTRUE(plot_lm)) geom_smooth(linewidth = 0.5, method=lm, se=FALSE) } +
        scale_color_viridis_d() +
        guides(color = guide_legend(title = 'cluster')) +
        lims(x = c(-abs_max, abs_max),
             y = c(-abs_max, abs_max)) +
        labs(y = 'ATAC log2(FC)', x = 'RNA log2(FC)')
    return(corr_line_plot)
}


diff_dat_filt <- diff_dat %>% 
  filter(pvalue_rna < alpha_level, pvalue_atac < alpha_level, 
         abs(log2FoldChange_atac) >= log2fc_thresh, abs(log2FoldChange_rna) >= log2fc_thresh)

cor_logfc <- diff_dat %>% 
    group_by(cluster_id) %>% 
    do(tidy(
        cor.test(.$log2FoldChange_atac, .$log2FoldChange_rna,
                 method = "spearman")
    ))
kable(cor_logfc)

lm_logfc <- diff_dat_filt %>% 
    group_by(cluster_id) %>% 
    do(tidy(lm(
        log2FoldChange_atac ~ log2FoldChange_rna, .
    )))


ggplotly(
  diff_dat_filt %>% plot_corr_line(plot_lm = TRUE), 
  tooltip = 'text'
 )
```

### Overlap significant genes

#### between clusters

```{r signif_genes_overlap}
gene_lists <- diff_dat_filt %>% 
    select(gene_name, cluster_id) %>% 
    arrange(gene_name) %>% 
    unstack()

Venn(gene_lists) %>% 
    process_data() %>% 
    plot_venn() +
    scale_fill_viridis_c() +
    theme(legend.position = 'none')
```

#### between expression-accessibility bins

```{r gene_quadrants}
diff_dat_filt_cat <- diff_dat_filt %>% 
    mutate(category = case_when(
        log2FoldChange_rna > 0 & log2FoldChange_atac > 0 ~ 'HE-HA',
        log2FoldChange_rna > 0 & log2FoldChange_atac < 0 ~ 'HE-LA',
        log2FoldChange_rna < 0 & log2FoldChange_atac > 0 ~ 'LE-HA',
        log2FoldChange_rna < 0 & log2FoldChange_atac < 0 ~ 'LE-LA',
        TRUE ~ NA_character_
    ))

diff_dat_filt_cat %>% 
    select(cluster_id, category, gene_name,
           starts_with('log2'), starts_with('pval')) %>% 
    arrange(cluster_id, category) %>% 
    write_csv(params$atac_rna_signif_genes_csv)

diff_dat_filt_cat %>% 
    select(gene_name, category) %>% 
    arrange(gene_name) %>% 
    unstack() %>% 
    Venn() %>% 
    process_data() %>% 
    plot_venn() +
    scale_fill_viridis_c() +
    theme(legend.position = 'none')

```

#### Overlap in expression & accessbility between clusters

- HE-HA: High Expression (RNAseq) and High Accessibility (ATACseq)
- LE-LA: Low Expression (RNAseq) and Low Accessibility (ATACseq)

```{r he-ha_genes, width='100%'}
plot_venn_categories <- function(dat, expr_category = 'HE-HA') {
  dat %>% 
    filter(category == expr_category) %>%
    select(gene_name, cluster_id) %>% 
    arrange(gene_name) %>% 
    unstack() %>% 
    Venn() %>% 
    process_data() %>% 
    plot_venn() +
    scale_fill_viridis_c() +
    labs(title = expr_category) +
    theme(legend.position = 'none')
}
categories <- diff_dat_filt_cat %>% pull(category) %>% unique() %>% sort()
plots <- map(categories, plot_venn_categories, dat = diff_dat_filt_cat)

plot_grid <- ggplot() +
    geom_hline(yintercept = 0,
               linetype = 'dashed',
               color = 'darkgrey') +
    geom_vline(xintercept = 0,
               linetype = 'dashed',
               color = 'darkgrey') +
    scale_x_continuous(breaks = c(0)) +
    scale_y_continuous(breaks = c(0)) +
    labs(x = 'Expression (RNA)', y = 'Accessibility (ATAC)') +
    theme(axis.line = element_line(arrow = grid::arrow()))
# use patchwork to compose single plot
plot_grid +
    inset_element(plots[[1]], left = 0.5, bottom = 0.5, right = 1, top = 1) + # upper right
    inset_element(plots[[3]], left = 0, bottom = 0.5, right = 0.5, top = 1) + # upper left
    inset_element(plots[[2]], left = 0.5, bottom = 0, right = 1, top = 0.5) + # lower right
    inset_element(plots[[4]], left = 0, bottom = 0, right = 0.5, top = 0.5)   # lower left
```

### Enrichment analysis

TODO run clusterProfiler separately on each cluster's HE-HA gene list. need > 100 genes for good ORA call

https://yulab-smu.top/biomedical-knowledge-mining-book/universal-api.html#msigdb-gsea

use category C2 CP (curated gene sets - canonical patheays )

## Infer TF-gene regulatory network


