---
title: 'integrated differential ATACseq and RNAseq'
date: today
engine: knitr
execute:
    message: false
    warning: false
format:
    html:
        self-contained: true
        toc: true
        code-fold: true
params:
  cpus: 1
  artifact_dir: NULL
  input_dir: NULL
  meta: NULL
  atac_rna_tsv:   "output/rowbind_logfc/concat.atac_rna.tsv"
  atac_rna_venn:  "output/join_atac_rna/joined_atac_rna.log"
  set_counts_tsv: "output/rowbind_count/concat.set_counts.tsv"
---

```{r setup}
#| message: false
library(broom)
library(dplyr)
library(ggplot2)
library(ggVennDiagram)
library(glue)
library(knitr)
library(plotly)
library(readr)
library(scales)
library(stringr)
library(tidyr)
theme_sovacool <- ggplot2::theme_bw() +
    ggplot2::theme(
      legend.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      legend.box.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      panel.spacing = unit(1, 'lines'),
      plot.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      strip.background = element_blank(),
      strip.placement = 'outside'
    )
theme_set(theme_sovacool)
```



```{r read}
#| message: false
dat <- read_tsv(params$atac_rna_tsv) %>% 
    mutate(cluster_id = as.factor(cluster_id))
```


## Differential ATAC & RNAseq 

### Overlapping genes

Differential ATACseq peaks were intersected with known gene promoter regions.
Peaks that overlap promoter regions were kept and joined with the corresponding
differential gene expression data from RNAseq.

Inspired by Fig. 4A from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9581296/figure/F4/>

```{r plot_venn}
#| message: false
v <- read_delim(params$atac_rna_venn, delim = ':\t', 
           col_names = c('set','genes')) %>% 
    pivot_wider(names_from = set, values_from = genes) %>% 
    unlist()

venn_dat <- list(
    RNAseq = identity_transformer(glue("1:{v[['RNAseq']]}")),
    ATACseq = identity_transformer(glue("{v[['RNAseq']]-v[['joined']]}:{v[['RNAseq']]+v[['ATACseq']]+v[['joined']]}"))
)

Venn(venn_dat) %>% 
    process_data(shape_id = "201") %>% 
    plot_venn() +
    scale_fill_distiller(palette = 'YlGnBu', direction = 1) +
    theme(legend.position = 'none')

```

Inspired by Fig. 3B-C from <https://www.science.org/doi/10.1126/science.abe1505#F3>

```{r plot_histogram}
overlap_counts <- read_tsv(params$set_counts_tsv) 

plot_hist <- function(dat) {
    dat %>%
        ggplot(aes(count, fill = count_type)) +
        geom_histogram(binwidth = 1) +
        facet_grid(cluster_id ~ count_type, switch = 'x') + #, strip.position = 'bottom') +
        labs(x = '') +
        theme(legend.position = 'none')
}

overlap_counts %>%
    filter(count > 0) %>% 
    plot_hist()

overlap_counts %>% 
    plot_hist() +
    scale_y_log10(labels = label_log(digits = 2))
```

### Correlation

Inspiration

- Fig. 4B from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9581296/figure/F4/>
- Fig. 2A-B from <https://epigeneticsandchromatin.biomedcentral.com/articles/10.1186/s13072-019-0260-2/figures/2>

```{r plot_corr_line}
cor_logfc <- dat %>% 
    group_by(cluster_id) %>% 
    do(tidy(cor.test(.$log2FoldChange_atac, .$log2FoldChange_rna, 
                     method = "spearman")))
kable(cor_logfc)
lm_logfc <- dat %>% group_by(cluster_id) %>% 
  do(tidy(lm(log2FoldChange_atac ~ log2FoldChange_rna, .)))

min_max <- list(
  min = ~min(.x, na.rm = TRUE), 
  max = ~max(.x, na.rm = TRUE)
)
abs_max <- dat %>% 
    summarize(across(starts_with('log2FoldChange'), min_max)) %>% 
    as.list() %>% unname() %>% unlist() %>% abs() %>% max()
corr_line_plot <- dat %>% 
    mutate(across(starts_with('log2FoldChange'), ~ round(., 2))) %>% 
    ggplot(aes(y = log2FoldChange_atac, x = log2FoldChange_rna,
               color = cluster_id, group = cluster_id, 
               text = glue('gene name:\t{gene_name}', 
                           'cluster:\t{cluster_id}',
                           'ATAC log2(FC):\t{log2FoldChange_atac}',
                           'RNA log2(FC):\t{log2FoldChange_rna}',
                           .sep = '\n')
               )
           ) +
    geom_hline(yintercept = 0, linetype = 'dashed', color = 'darkgrey') +
    geom_vline(xintercept = 0, linetype = 'dashed', color = 'darkgrey') +
    geom_point(alpha = 0.7) +
    #geom_smooth(method=lm, se=FALSE) +
    scale_color_viridis_d() +
    guides(color = guide_legend(title = 'cluster')) +
    lims(x = c(-abs_max, abs_max), y = c(-abs_max, abs_max)) +
    labs(y = 'ATAC log2(FC)', x = 'RNA log2(FC)')
ggplotly(corr_line_plot, tooltip = 'text')
```


## Infer TF-gene regulatory network


