---
title: Integrated ATAC-seq and RNA-seq for SCLC
author: CCBR
date: today
format: 
    revealjs:
        theme: [default, custom.scss]
        embed-resources: true
params:
  atac_rna_tsv:   "output/rowbind_logfc/concat.atac_rna.tsv"
  atac_rna_venn:  "output/join_atac_rna/joined_atac_rna.log"
  set_counts_tsv: "output/rowbind_count/concat.set_counts.tsv"
---


```{r setup}
#| message: false
library(broom)
library(dplyr)
library(ggplot2)
library(ggVennDiagram)
library(glue)
library(knitr)
library(patchwork)
library(plotly)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tidyr)
theme_sovacool <- ggplot2::theme_bw() +
    ggplot2::theme(
      legend.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      legend.box.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      panel.spacing = unit(1, 'lines'),
      plot.margin = ggplot2::margin(0, 0, 0, 0, unit = "pt"),
      strip.background = element_blank(),
      strip.placement = 'outside'
    )
theme_set(theme_sovacool)
```

## Differential ATACseq & RNAseq 

<!--
Inspired by Fig. 4A from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9581296/figure/F4/>
-->

```{r plot_venn}
#| message: false
v <- read_delim(params$atac_rna_venn, delim = ':\t', 
           col_names = c('set','genes')) %>% 
    pivot_wider(names_from = set, values_from = genes) %>% 
    unlist()

venn_dat <- list(
    RNAseq = identity_transformer(glue("1:{v[['RNAseq']]}")),
    ATACseq = identity_transformer(glue("{v[['RNAseq']]-v[['joined']]}:{v[['RNAseq']]+v[['ATACseq']]+v[['joined']]}"))
)

Venn(venn_dat) %>% 
    process_data(shape_id = "201") %>% 
    plot_venn() +
    scale_fill_distiller(palette = 'YlGnBu', direction = 1) +
    theme(legend.position = 'none')

```

ATACseq peaks in known promoter regions intersected with differential genes from RNAseq.


## Peaks in genes & genes in peaks
<!--
Inspired by Fig. 3B-C from <https://www.science.org/doi/10.1126/science.abe1505#F3>
-->
```{r peaks_in_genes}
overlap_counts <- read_tsv(params$set_counts_tsv) 

plot_hist <- function(dat) {
    dat %>%
        ggplot(aes(count, fill = count_type)) +
        geom_histogram(binwidth = 1) +
        facet_grid(cluster_id ~ count_type, switch = 'x') + #, strip.position = 'bottom') +
        labs(x = '') +
        theme(legend.position = 'none')
}


overlap_counts %>% 
    plot_hist() +
    scale_y_continuous(labels = label_comma())

#overlap_counts %>% plot_hist() + scale_y_log10(labels = label_log(digits = 2))
```

## Peaks in genes & genes in peaks

```{r peaks_in_genes_filt}
overlap_counts %>%
    filter(count > 0) %>% 
    plot_hist()
```

## Differential ATAC & RNA-seq Correlation

<!--
Inspiration

- Fig. 4B from <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9581296/figure/F4/>
- Fig. 2A-B from <https://epigeneticsandchromatin.biomedcentral.com/articles/10.1186/s13072-019-0260-2/figures/2>
-->

```{r read}
#| message: false
diff_dat <- read_tsv(params$atac_rna_tsv) %>% 
    mutate(cluster_id = as.factor(cluster_id))
```


```{r corr_all_genes, message=FALSE}
alpha_level <- 0.05
log2fc_thresh <- 0 # setting to zero removes filtering by log2FC

cor_logfc <-  cor.test(diff_dat$log2FoldChange_atac, 
         diff_dat$log2FoldChange_rna,
         method = "spearman") %>% 
    tidy()
#cor_logfc %>% kable()

lm_logfc <- diff_dat %>%
    lm(log2FoldChange_atac ~ log2FoldChange_rna, .) 
lm_rsq <- summary(lm_logfc)$r.squared

abs_max <- 5.004292 # from filtered data below. this will exclude some points from the unfiltered data.
    
diff_dat %>%
    mutate(across(starts_with('log2FoldChange'), ~ round(., 2))) %>%
    mutate(signif = case_when(
        pvalue_atac < alpha_level & pvalue_rna < alpha_level ~ TRUE,
        TRUE ~ FALSE)
    ) %>%
    dplyr::arrange(signif) %>% # plot true on top of false
    ggplot(
        aes(
            y = log2FoldChange_atac,
            x = log2FoldChange_rna
        )
    ) +
    geom_hline(
        yintercept = 0,
        linetype = 'dashed',
        color = 'darkgrey'
    ) +
    geom_vline(
        xintercept = 0,
        linetype = 'dashed',
        color = 'darkgrey'
    ) +
    geom_point(aes(fill = signif), alpha = 0.7, shape = 21, stroke = 0) +
    geom_smooth(method = lm, 
                linewidth = 0.5,
                se = FALSE, show.legend = FALSE) +
    scale_fill_manual(values = c(`TRUE` = "#252525", `FALSE` = "#D9D9D9")) +
    guides(fill = guide_legend(title = glue('ATAC & RNA\np-values < {alpha_level}'))) +
    lims(x = c(-abs_max, abs_max),
         y = c(-abs_max, abs_max)) +
    labs(y = 'ATAC log2(FC)', x = 'RNA log2(FC)')
```

Spearman's rho: `r cor_logfc$estimate` (p-value < 2.2e-16)


## Correlation within clusters

```{r}

cor_logfc <- diff_dat %>% 
    group_by(cluster_id) %>% 
    do(tidy(
        cor.test(.$log2FoldChange_atac, .$log2FoldChange_rna,
                 method = "spearman")
    )) %>% 
    select(cluster_id, estimate, p.value) %>% 
    rename(cluster = cluster_id, `Spearman's Rho` = estimate, `p-value` = p.value) %>% 
    mutate(`p-value` = case_when(`p-value` < 0.05 ~ '< 0.05',
                               TRUE ~ '>= 0.05'))
kable(cor_logfc)
```


## Differential ATAC & RNA-seq Correlation


```{r corr_signif_genes, message=FALSE, warning=FALSE, out.width='80%'}

min_max <- list(min = ~ min(.x, na.rm = TRUE),
                max = ~ max(.x, na.rm = TRUE))
plot_corr_line <- function(dat, plot_lm = FALSE) {
    
    abs_max <- dat %>%
        summarize(across(starts_with('log2FoldChange'), min_max)) %>%
        as.list() %>% unname() %>% unlist() %>% abs() %>% max()
    corr_line_plot <- dat %>%
        mutate(across(starts_with('log2FoldChange'), ~ round(., 2))) %>%
        ggplot(
            aes(
                y = log2FoldChange_atac,
                x = log2FoldChange_rna,
                color = cluster_id,
                group = cluster_id,
                text = glue(
                    'gene name:\t{gene_name}',
                    'cluster:\t{cluster_id}',
                    'ATAC log2(FC):\t{log2FoldChange_atac}',
                    'RNA log2(FC):\t{log2FoldChange_rna}',
                    .sep = '\n'
                )
            )
        ) +
        geom_hline(
            yintercept = 0,
            linetype = 'dashed',
            color = 'darkgrey'
        ) +
        geom_vline(
            xintercept = 0,
            linetype = 'dashed',
            color = 'darkgrey'
        ) +
        geom_point(alpha = 0.5) +
        {if (isTRUE(plot_lm)) geom_smooth(linewidth = 0.5, method=lm, se=FALSE) } +
        scale_color_viridis_d() +
        guides(color = guide_legend(title = 'cluster')) +
        lims(x = c(-abs_max, abs_max),
             y = c(-abs_max, abs_max)) +
        labs(y = 'ATAC log2(FC)', x = 'RNA log2(FC)')
    return(corr_line_plot)
}


diff_dat_filt <- diff_dat %>% 
  filter(pvalue_rna < alpha_level, pvalue_atac < alpha_level, 
         abs(log2FoldChange_atac) >= log2fc_thresh, abs(log2FoldChange_rna) >= log2fc_thresh)


lm_logfc <- diff_dat_filt %>% 
    group_by(cluster_id) %>% 
    do(tidy(lm(
        log2FoldChange_atac ~ log2FoldChange_rna, .
    )))


ggplotly(
  diff_dat_filt %>% plot_corr_line(plot_lm = TRUE), 
  tooltip = 'text'
 )
```

Genes filtered to p-value < `r alpha_level` for both ATACseq & RNAseq log2FC

<!--
## Overlap between clusters

```{r signif_genes_overlap}
gene_lists <- diff_dat_filt %>% 
    select(gene_name, cluster_id) %>% 
    arrange(gene_name) %>% 
    unstack()

Venn(gene_lists) %>% 
    process_data() %>% 
    plot_venn() +
    scale_fill_viridis_c() +
    theme(legend.position = 'none')
```
-->

## Expression & accessbility

```{r gene_quadrants, out.width='80%'}
diff_dat_filt_cat <- diff_dat_filt %>% 
    mutate(category = case_when(
        log2FoldChange_rna > 0 & log2FoldChange_atac > 0 ~ 'HE-HA',
        log2FoldChange_rna > 0 & log2FoldChange_atac < 0 ~ 'HE-LA',
        log2FoldChange_rna < 0 & log2FoldChange_atac > 0 ~ 'LE-HA',
        log2FoldChange_rna < 0 & log2FoldChange_atac < 0 ~ 'LE-LA',
        TRUE ~ NA_character_
    ))

p <- diff_dat_filt_cat %>%
        mutate(across(starts_with('log2FoldChange'), ~ round(., 2))) %>%
        ggplot(
            aes(
                y = log2FoldChange_atac,
                x = log2FoldChange_rna,
                color = category,
                group = category,
                text = glue(
                    'gene name:\t{gene_name}',
                    'cluster:\t{cluster_id}',
                    'ATAC log2(FC):\t{log2FoldChange_atac}',
                    'RNA log2(FC):\t{log2FoldChange_rna}',
                    .sep = '\n'
                )
            )
        ) +
        geom_hline(
            yintercept = 0,
            linetype = 'dashed',
            color = 'darkgrey'
        ) +
        geom_vline(
            xintercept = 0,
            linetype = 'dashed',
            color = 'darkgrey'
        ) +
        geom_point(alpha = 0.5) +
        scale_color_viridis_d() +
        guides(color = guide_legend(title = '')) +
        lims(x = c(-abs_max, abs_max),
             y = c(-abs_max, abs_max)) +
        labs(y = 'ATAC log2(FC)', x = 'RNA log2(FC)')

ggplotly(p, tooltip = 'text')
```

- HE-HA: High Expression (RNAseq) and High Accessibility (ATACseq)
- LE-LA: Low Expression (RNAseq) and Low Accessibility (ATACseq)

## Expression & accessbility

```{r he-ha_genes, width='100%'}
plot_venn_categories <- function(dat, expr_category = 'HE-HA') {
  dat %>% 
    filter(category == expr_category) %>%
    select(gene_name, cluster_id) %>% 
    arrange(gene_name) %>% 
    unstack() %>% 
    Venn() %>% 
    process_data() %>% 
    plot_venn() +
    scale_fill_viridis_c() +
    labs(title = expr_category) +
    theme(legend.position = 'none')
}
categories <- diff_dat_filt_cat %>% pull(category) %>% unique() %>% sort()
plots <- map(categories, plot_venn_categories, dat = diff_dat_filt_cat)

plot_grid <- ggplot() +
    geom_hline(yintercept = 0,
               linetype = 'dashed',
               color = 'darkgrey') +
    geom_vline(xintercept = 0,
               linetype = 'dashed',
               color = 'darkgrey') +
    scale_x_continuous(breaks = c(0)) +
    scale_y_continuous(breaks = c(0)) +
    labs(x = 'Expression (RNA)', y = 'Accessibility (ATAC)') +
    theme(axis.line = element_line(arrow = grid::arrow()))
# use patchwork to compose single plot
plot_grid +
    inset_element(plots[[1]], left = 0.5, bottom = 0.5, right = 1, top = 1) + # upper right
    inset_element(plots[[3]], left = 0, bottom = 0.5, right = 0.5, top = 1) + # upper left
    inset_element(plots[[2]], left = 0.5, bottom = 0, right = 1, top = 0.5) + # lower right
    inset_element(plots[[4]], left = 0, bottom = 0, right = 0.5, top = 0.5)   # lower left
```


## In Progress & Next Steps

- Enrichment analysis on high-expression high-accessibility genes (clusterProfiler)
- Construct TF-gene regulatory network (TOBIAS)
- Analyze variability in chromatin accessibility (chromVar)



